---
title: "NMF Tutorial"
author: "Helio"
date: "25/04/2022"
output: html_document
---

# load libraries
```{r}
library(NMF)
# library(bio)
# install.packages('biobase')
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("Biobase")
library(tidyverse)

```


# Load datasets


gene expression data from Golub et al. (1999).
Molecular Classification of Cancer: Class Discovery and Class
Prediction by Gene Expression Monitoring"

- 27 patients with acute lymphoblastic leukemia (ALL) 
-  11 patients with acute myeloid leukemia (AML).
- columns  - patient based on gene expression
- rows 


```{r}
library(NMF)
data(esGolub) 
# see details of the dataset
?esGolub
esGolub <- esGolub[1:200,] 

# get a subset of data
esGolub1<- as.data.frame(esGolub@assayData[["exprs"]])[1:200,]

esGolub1

# learn about the function

?NMF::nmf

# store name of genes
esGolub1$gene_names<- row.names(esGolub1)

```




Step 1 - INITIALIZATION 
- Specifying the seeding method

```{r}

NMF::nmfSeed()
# [1] "none"   "random" "ica"    "nndsvd"

# single run and single numeric seed for reproducibility
res_1run_1seed <- nmf(esGolub1[,1:38], 3, seed=123456) 

summary(summary(res2))

res_100run_1seed

```
# Multiple runs for stohcastic seed
# default keep the best fit over all run
# typically 100 - 200 runs

```{r}
library(NMF)
res_100run_1seed <- NMF::nmf(esGolub1[,1:38], 3, nrun = 100,
                             .options = 'v')
options(scipen = 999)

summary(res_100run_1seed)

# keep all
res_100run_1seed_all <- NMF::nmf(esGolub1[,1:38], 3, nrun = 100,
                             .options = list(keep.all = TRUE, 'v'))

View(res_100run_1seed_all[[1]])
View(res_100run_1seed_all[[2]])
```


# HOW MANY COMPONENTES = K = ?
we will do 100 initialization for each k  = from 2 to 6 
use seed for reproducibility

```{r}
res_k <-NMF::nmf(esGolub1[,1:38], 2:6, nrun=100, seed=123456,
                 .options = 'v')
# takes a bit
# verbose
# Compute NMF rank= 2  ... + measures ... OK
# Compute NMF rank= 3  ... + measures ... OK
# Compute NMF rank= 4  ... + measures ... OK
# Compute NMF rank= 5  ... + measures ... OK
# Compute NMF rank= 6  ... + measures ... OK


res_k

# we can plot a summary of quality measures

NMF::plot(res_k) #+ geom_vline(xintercept = 3, linetype = "dashed") 
# do 3 then 4

```


Consensus map

```{r}
# and the function consensusmap generates  heatmaps  of  the  consensus  matrix  for  each  value  of  the  rank.
?consensusmap
consensusmap(res_k, annCol=esGolub, labCol=NA, labRow=NA)

```

SHUFFLING to avoid overfitting to noise

```{r}
# shuffle original data 

V.random <- randomize(esGolub1[,1:38]) 

# estimate quality measures from the shuffled data (use default NMF algorithm) 
estim.k.random <- nmf(V.random, 2:6, nrun=100, seed=123456,
                      .option = 'v') 

# Compute NMF rank= 2  ... + measures ... OK
# Compute NMF rank= 3  ... + measures ... OK
# Compute NMF rank= 4  ... + measures ... OK
# Compute NMF rank= 5  ... + measures ... OK
# Compute NMF rank= 6  ... + measures ... OK

# then we can assess the quality of random estimation to our estimation

# plot measures on same graph (x, y) 
plot(res_k, estim.k.random) + geom_vline(xintercept = 3)


```

We select values of k where the magnitude of the cophenetic correlation coefficient begins to fall (see below).
so now we can fit the final model


```{r}
res_k3 <-NMF::nmf(esGolub1[,1:38], r = 3, nrun = 100, seed=123456,
                 .options = list( 'v','t')) #error tracking

res_k3 <-NMF::nmf(esGolub1[,1:38], r = 3, nrun = 100, seed=123456,
                 .options = 't') #error tracking
plot(res_k3)
```

plotting results


```{r}

layout(cbind(1,2)) 

# basis components 
basismap(res_k3, subsetRow=TRUE) # weights/ amplitudes
# mixture coefficients 
coefmap(res_k3) #hidden variables

# access moe info
# res_k3

```

NMF on facial action activation

load face data

```{r}

db_of_morph_study <- readRDS("~/Library/CloudStorage/OneDrive-UniversityofBirmingham/Lab meetings/NMF_tutorial/db_of_morph_study.Rds")

View(db_of_morph_study)

# db_of_morph_study$morph
db_of_morph_study$Dataset
db_of_morph_study %>%
  subset(morph == "Original" & Dataset == "ADFES")%>%
  ggplot(aes(timebin))+
  geom_histogram()

table(db_of_morph_study$timebin)

# select only AU columns and metadata
colnames(db_of_morph_study)
unique(db_of_morph_study$AU01_c)

db_of<- db_of_morph_study[, c(1:12,58:74)]

# rename the AUS into more descriptivinfotheo::

colnames(db_of)

colnames(db_of)
# separate morphs and not morphs
db_of_morph<- subset(db_of, morph == "Morph")
db_of_natural<- subset(db_of, morph == "Original")

```

we can start with an NMF that ignores the time component
```{r}

# load AU wiki
AU_Wiki <- readxl::read_excel("AU_Wiki.xlsx")
View(AU_Wiki)

AU_Wiki$AU_Code_OF<- paste0("AU", paste0(AU_Wiki$AU_CODE), paste0("_r"))

colnames(db_of_natural)
AU_OF_codes<- colnames(db_of_natural[,13:29])

as.data.frame(AU_OF_codes)

AU_Wiki$AU_OF_codes<-AU_Wiki$AU_Code_OF

AU_OF_codes_merged <- left_join(as.data.frame(AU_OF_codes), AU_Wiki)

colnames(db_of_natural)


# put underscores to avoid issues later
AU_OF_codes_merged$FACS_name <- chartr(" ", "_", AU_OF_codes_merged$FACS_name)

AU_OF_transposed<- t(AU_OF_codes_merged$FACS_name)

```


NMF - no temporal stuff
```{r}
db_of_natural1<- db_of_natural
# change names to descriptive AUS
AU_OF_transposed_names <-  c(paste0(colnames(db_of_natural[,1:12])),(AU_OF_transposed[1,]))

names(db_of_natural1)<- AU_OF_transposed_names

colnames(db_of_natural1)

db_of_natural1_agg<- db_of_natural1[,c(1,4,5,13:29)]%>%
  group_by(filename, Emotion, Dataset)%>%
  summarise_if(is.numeric, mean, na.rm = TRUE)

db_of_natural1_agg

```

inotialise and find k
```{r}
colnames(db_of_natural1_agg)
unique(db_of_natural1_agg$Emotion)
# single run and single numeric seed for reproducibility

# db_of_natural1_agg[,2:19]+.001
res_find_k <- nmf(db_of_natural1_agg[,3:19], 2:6, 
                      seed=123456,
                      nrun = 100,
                       .options = 'v') 

summary(res_find_k)

NMF::plot(res_find_k)+
  geom_vline(xintercept = 3)

?NMF::consensusmap
NMF::consensusmap(res_find_k)


```
randomize

```{r}
V.random_em <- randomize(db_of_natural1_agg[,3:19]) 

# estimate quality measures from the shuffled data (use default NMF algorithm) 
estim.k.random_em <- nmf(V.random_em, 2:6, nrun=100, seed=123456,
                      .option = 'v') 

# Compute NMF rank= 2  ... + measures ... OK
# Compute NMF rank= 3  ... + measures ... OK
# Compute NMF rank= 4  ... + measures ... OK
# Compute NMF rank= 5  ... + measures ... OK
# Compute NMF rank= 6  ... + measures ... OK

# then we can assess the quality of random estimation to our estimation

# plot measures on same graph (x, y) 
plot(res_find_k, estim.k.random_em)+geom_vline(xintercept = 3)


```

```{r}
res_k3_em <-NMF::nmf(db_of_natural1_agg[,4:19], r = 3, nrun = 100, seed=123456)


layout(cbind(1,2)) 

summary(res_k3_em)

# basis components 
NMF::basismap(res_k3_em) # weights/ amplitudes
# mixture coefficients 
coefmap(res_k3_em) #hidden variables

# access moe info
# res_k3_em


db_of_natural1_agg$NMFtable<- as.data.frame(res_k3_em@fit@W)
db_of_natural1_agg$comp1 = db_of_natural1_agg$NMFtable$V1
db_of_natural1_agg$comp2 = db_of_natural1_agg$NMFtable$V2

db_of_natural1_agg$comp3 = db_of_natural1_agg$NMFtable$V3

colnames(db_of_natural1_agg)

library(viridis)

db_of_natural1_agg[,c(2,3,22:24)] %>% 
  gather(component, coef, -Emotion, -Dataset)%>%
  group_by(Emotion, component,Dataset)%>%
  summarise_if(is.numeric, mean, na.rm = TRUE)%>%
  ggplot(aes(Emotion, component, fill = coef))+
  geom_tile()+
  scale_fill_viridis(option="magma")+
  theme_classic()+
  facet_grid(~Dataset)



db_of_natural1_agg[,c(2,3,22:24)] %>% 
  gather(component, coef, -Emotion, -Dataset)%>%
  group_by(Emotion, component,Dataset)%>%
  ggplot(aes(Emotion, coef, , color = Emotion, fill = Emotion))+
  geom_jitter(alpha = .1, position = position_dodge(1))+
  stat_summary(geom = "pointrange")+
  geom_boxplot(alpha = .1)+
    theme_classic()+
  facet_grid(component~Dataset)+
  ggpubr::stat_compare_means()+
  scale_color_brewer(palette = "Dark2")+
  scale_fill_brewer(palette = "Dark2")


db_of_natural1_agg[,c(2,3,22:24)] %>% 
  gather(component, coef, -Emotion, -Dataset)%>%
  group_by( component,Dataset)%>%
  ggplot(aes(Dataset, coef , color = Dataset, fill = Dataset))+
  geom_jitter(alpha = .1, position = position_dodge(1))+
  stat_summary(geom = "pointrange")+
  geom_boxplot(alpha = .1)+
    theme_classic()+
  facet_grid(~component)+
  ggpubr::stat_compare_means()+
  scale_color_brewer(palette = "Dark2")+
  scale_fill_brewer(palette = "Dark2")

# we can do clasification with this
db_of_natural1_agg

write_csv(db_of_natural1_agg, "db_of_natural1_agg.csv")
```


<!-- now for the temporal -->


```{r}
colnames(db_of_natural1)
db_of_natural1$timebin
table(is.na(db_of_natural1))
res_k3_em_ts <-NMF::nmf(db_of_natural1[,13:29], r = 3, nrun = 100, seed=123456,
                        .options = 'v')

summary(res_k3_em_ts)

NMF::coefmap(res_k3_em_ts)
NMF::basismap(res_k3_em_ts)


db_of_natural1$NMFtable<- as.data.frame(res_k3_em_ts@fit@W)
db_of_natural1$comp1 = db_of_natural1$NMFtable$V1
db_of_natural1$comp2 = db_of_natural1$NMFtable$V2
db_of_natural1$comp3 = db_of_natural1$NMFtable$V3


# morph
colnames(db_of_morph)<- c(AU_OF_transposed_names,"NMFtable", "comp1", "comp2",
                          "comp3")

db_of_morph1<- 
res_k3_em_ts_morph <-NMF::nmf(db_of_morph[,13:29], r = 3, nrun = 100, seed=123456,
                        .options = 'v')

summary(res_k3_em_ts_morph)

NMF::coefmap(res_k3_em_ts_morph)
NMF::basismap(res_k3_em_ts_morph)


db_of_morph$NMFtable<- as.data.frame(res_k3_em_ts_morph@fit@W)
db_of_morph$comp1 = db_of_morph$NMFtable$V1
table(is.na(db_of_morph$comp1))
db_of_morph$comp2 = db_of_morph$NMFtable$V2
db_of_morph$comp3 = db_of_morph$NMFtable$V3

colnames(db_of_natural1)
bind_rows(db_of_natural1, db_of_morph)[,c(2:5,31:33)]%>%
  gather(component, coef, -timebin,-Emotion, -Dataset, -morph)%>%
  # group_by(component,Dataset)%>%
  # ggplot(aes(Dataset, coef , color = Dataset, fill = Dataset))+
  
  ggplot(aes(timebin, coef, color = component))+
  geom_smooth()+
  facet_grid(Dataset~Emotion*morph)+
  theme_classic()

colnames(db_of_morph)

bind_rows(db_of_natural1, db_of_morph)[,c(1,2:5,31:33)]%>%
  # gather( component, coef, -timebin,-Emotion, -Dataset, -morph, -filename)%>%
  group_by(filename,timebin, Emotion, Dataset, morph)%>%
  summarise_at(c("comp1", "comp2", "comp3"), mean, na.rm = TRUE)

nmf_agg_ts <- bind_rows(db_of_natural1, db_of_morph)[,c(1,2:5,31:33)]%>%
  # gather( component, coef, -timebin,-Emotion, -Dataset, -morph, -filename)%>%
  group_by(filename, Emotion, Dataset, morph)%>%
  summarise_at(c("comp1", "comp2", "comp3"), mean, na.rm = TRUE)
```

